Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-09 demystify.tex
updates

For example, in the 2024 admission cycle, there is a huge surge in students interested in AI and NLP (thanks, ChatGPT!). Thus, AI/NLP faculty might be overwhelm with potential students and cannot take additional students, many of whom have excellent profiles.

@@

Even if you do not have research experience, you can talk about your personal projects, as long as they can help show you can do research. For example, if you have an open-source project on Github that is used by many people, has lots of stars in Github, do talk about it. If you write technical, research-like blogs, talk about them too.

@@

It is understandable that you might feel uncomfortable disclosing this, fearing you would lost a chance of working with a faculty who might have a bias.  However, if someone has this issue, then you might not want to work with them anyway (and such bias is not acceptable in academic and likely would not be tolerated by the university). In my experience, it is the opposite: \emph{many profs. actively seek out to work with students from diverse backgrounds and view diversity as an asset to their research group}.

@@

@@

\section{How much do \emph{you} cost?}\label{sec:ra-cost}
PhD students often ask why their salary is so low compared to ludicrous grants their advisers get. They also wonder why their offer letters sometimes show that their benefits higher than what they actually receive (i.e., stipend).  This section aims to shed some light to these questions.

@@

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$29000 and a 3-month summer of \$10000 (about a third of the 9-month stipend). I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., MIT charges around \$50K) and a 58.9\% rate on \emph{indirect cost}, which is a typical rate charged for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

@@

\section{Buying Computer Equipment}\label{sec:buying-equipment} Many students get very excited about getting to buy new computer equipment and electronics when they start their PhD. However, you should check with your adviser first.  Many profs. have funding to buy you a computer and other equipments (e.g., keyboards, headphones, tablets, books, etc).

\section{Can I negotiate my PhD offer?}\label{sec:negotiate}
For stipend, unlikely, as TA is set by the university and while RA is set by the adviser, it is often based on the department's standard, and it also would be unfair to other students if you get more. However, your stipend is often increased every year (by a small amount and automatically) and will likely increase after you pass your qualifying exam.

For start date or TA assignment (e.g., TA'ing a specific course), it is uncommon but you can ask for it. Also, there is typically no moving allowance for PhD students. In short, things that are standard and set by the university or department are unlikely to be negotiable.  However, you can ask for things such as computer equipments (\autoref{sec:buying-equipment}), books, (and sometimes even furniture) and conference travel budget.

@@

In short, if you are not serious about an offer for any reason (e.g., still waiting for other offers), do not accept it.
If you are not sure, ask for more time to decide.  But do not accept an offer and then later reject it.  This can be considered \emph{rude and unprofessional}, and will likely burn bridges with that prof.

@@

\paragraph{Declining an offer} If you decide to decline or reject an offer, you can say: \emph{``Thank you for the offer. However, I have decided to accept another offer.  I appreciate your time and consideration.  I hope we can work together in the future.''}  Professors will understand and wish you good luck.  If you decide to reject an offer, do so quickly.

@@

Plagiarism and cheating (e.g., exams, assignments) is a BIG no-no in the US.  If you're caught cheating, you will face very heavy consequence and likely be expelled from the university (e.g., after the second time at GMU).   This is quite different from many international countries where cheating is common and often tolerated.  Faculty is extremely good at detecting cheating (we have been dealing with these situations so many times over so many years), and \emph{will} report cheating cases.  In short, whatever you do, don't cheat---not worth it.

@@

\section{Illegal Software} Using illegal/cracked software is very common in many countries (and even in the US). However, \emph{do not} install or use them on university computers, even those given to you from your adviser.  It is unlikely that the university will track you down, but it is the \emph{software company} that will.  They have very sophisticated tools to detect illegal software and will sue your university/department.  Imagine your department or adviser being sued for a large sum of money, and it is \emph{you} who caused it.  If you need to purchase software,  ask your adviser or the department.

>>>>

For example, in the 2024 admission cycle, there is a huge surge in students interested in AI and NLP (thanks, ChatGPT!). Thus, AI/NLP faculty might be 
overwhelmed and cannot consider many, even those with excellent profiles.

@@

Even if you do not have research experience, you can talk about your personal projects, as long as they can help show you can do research. For example, if you have an open-source project on Github that is used by many people, has lots of stars in Github, do talk about it. If you often write technical, research-like blogs with many viewers, talk about them too.

@@

It is understandable that you might feel uncomfortable disclosing this, fearing you would lost a chance of working with a faculty due to their bias.  However, if someone has this issue, then you might not want to work with them anyway (and such bias is not acceptable in academic and likely would not be tolerated by the university). In my experience, it is the opposite: \emph{many profs. actively seek out to work with students from diverse backgrounds and view diversity as an asset to their research group}.

@@

%https://www.edsurge.com/news/2018-05-18-how-cornell-university-diversified-its-incoming-phd-computer-science-student-body

%https://academia.stackexchange.com/questions/130591/is-it-standard-for-us-based-universities-to-consider-the-ethnicity-of-an-applica

%https://www.cs.cornell.edu/~bindel/paper/diversity.pdf

%https://cs.brown.edu/degrees/doctoral/applications/helpful-resources-applying-computer-science-phd-programs/

@@

\section{How much do \emph{I} cost?}\label{sec:ra-cost}
PhD students often ask why their salary is so low compared to ludicrous grants their advisers get. They also wonder why their offer letters sometimes show that their benefits higher than what they actually receive (i.e., stipend).

@@

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$29000 and a 3-month summer of \$10000 (about a third of the 9-month stipend). I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., Univ. of Chicago is a whooping \$70K) and a 58.9\% rate on \emph{indirect cost}, which is a typical rate charged for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

@@

\section{Buying Computer Equipment}\label{sec:buying-equipment} Many students get very excited about getting to buy new computer equipment and electronics when they start their PhD. However, you should check with your adviser first.  Many profs. have funding to buy you computers and other equipments (e.g., software, books, keyboards, headphones, tablets, etc).

\section{Can I negotiate my PhD offer?}\label{sec:negotiate}
For stipend, unlikely, as TA is set by the university and while RA is set by the adviser, it is often based on the department's standard, and it also would be unfair to other students if you get more. However, your stipend is often increased every year (automatically by a small amount) and also after you pass your qualifying exam.

For start date or TA assignment (e.g., TA'ing a specific course), it is uncommon but you can ask for it. Also, there is typically no moving allowance for PhD students. In short, standard things set by the university or department are unlikely non-negotiable.  However, you can ask for things such as computer equipments (\autoref{sec:buying-equipment}), books, (and sometimes even furniture) and conference travel budget.

@@

In short, if you are not serious about an offer for any reason (e.g., still waiting for other offers), \emph{do not accept it}.
If you are not sure, ask for more time to decide.  But do not accept an offer and then later reject it.  This can be considered \emph{rude and unprofessional}, and will likely burn bridges with that prof.

@@

\paragraph{Declining an offer} If you decide to decline or reject an offer, you can say: \emph{``Thank you for the offer. However, I have decided to accept another offer.  I appreciate your time and consideration.  I hope we can work together in the future.''}  Professors will understand and wish you luck.  If you decide to reject an offer, do so quickly.

@@

Plagiarism and cheating (e.g., exams, assignments) is a BIG no-no in the US.  If you're caught cheating, you will face serious consequence and likely be expelled from the university (e.g., after the second time at GMU).   This is quite different from many international countries where cheating is common and often tolerated.  Faculty is extremely good at detecting cheating (we have been dealing with these situations so many times over so many years), and \emph{will} report cheating cases.  In short, whatever you do, don't cheat---not worth it.

@@

\section{Illegal Software} Using illegal/cracked software is very common in many countries (and even in the US). However, \emph{do not} install or use them on university computers, even those given to you from your adviser.  It is unlikely that the university will track you down, but it is the \emph{software company} that will.  They have very sophisticated tools to detect illegal software and will sue your university/department.  Imagine your department or adviser being sued for a large sum of money, and it is \emph{you} who caused it.  If you need to purchase software,  ask your adviser or the department (\autoref{sec:buying-equipment}).
